`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation

### 2022-06-21 inst/paper/paper.Rmd
update-rmd-paragraphs

---
title: 'gosset: An R package for analysis and synthesis of ranking data in agricultural
  experimentation'
abstract: null
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

Participatory experimentation approaches have been increasingly applied in agricultural
research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general
agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA].
Recently developed approaches for on-farm experimentation,
such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019].
On the other hand, newly proposed approaches for synthesis of crop variety evaluation data 
largely depend on the analysis of ranking data [@brown2020data].
The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce  [@luce_individual_1959;@Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and 
Plackett-Luce models are available in R with the packages BradleyTerry2 and PlackettLuce
respectively [@BradleyTerry2; @Turner2020]. However, extended functionality was required for the entire 
data science workflow, which usually includes: (1) Data preparation and cleaning, 
(2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for 
converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2
respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the 
case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

The R package gosset provides functionality supporting the analysis
workflows in agricultural experimentation, especially with rank-based
approaches. The package is available in The Comprehensive R Archive Network (CRAN)
and can be installed by executing ``install.packages("gosset")``. The package is named 
in honor to William Sealy Gosset, known by the pen name ‘Student’, a pioneer of modern statistics
in small sample experimental design and analysis [@Ziliak_2019].

@@

**Data management and preparation**

*   **``rank_binomial``** transforms a ranking object into a binary comparisons, as required
by package BradleyTerry2 [@BradleyTerry2].  
*   **``rank_numeric``** converts numeric values into rankings. The parameter ``ascending``
indicates if the rankings should be made considering the numeric values in ascending order. 
The default is ``asceding = FALSE``. This function is useful when the data were collected 
as numerical observations, for instance, in a experiment measuring crop yield.  
*   **``rank_tricot``** transforms data in tricot format into rankings.

**Modelling**

*   **``AIC``** computes the Akaike Information Criterion @akaike1974] for a Bradley-Terry model fitted with
BradleyTerry2 [@BradleyTerry2] or a Plackett-Luce model fitted with [@Turner2020].  
*   **``btpermute``** deviance-based forward variable selection [@lysen2009permuted] procedure for Bradley-Terry models.  
*   **``compare``** compare the agreenment between methods.  
*   **``crossvalidation``** performs k-fold cross-validation, where k could be specified by the user.The default is
10-fold. Folds can be provided as a vector for a custom cross-validation, such as blocked cross-validation.
*   **``forward``** executes forward variable selection with cross-validation.  
*   **``kendallTau``** computes the Kendall-tau rank correlation coefficient between two rankings [@kendall_1938].  
*   **``kendallW``** computes Kendall's W (coefficient of concordance) among observed rankings and those predicted by
the Plackett-Luce model.  
*   **``pseudoR2``** computes goodness-of-fit measure McFadden's pseudo-R2 [@mcfadden1973conditional].

**Visualization and results presentation**

*   **``plot``** provides an alternative to the default plot function of the PlackettLuce package.

*   **``regret``**

*   **``reliability``**

*   **``worth_bar``**

*   **``worth_map``**

# Illustrative Examples
To demonstrate the functionality of the gosset package, we use the `nicabean` dataset,
which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.)
varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot 
approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties
of common bean. The varieties were randomly assigned as incomplete blocks of size three 
(out of 10 varieties). The farmers assessed which of the three varieties has the best and
worst performance in nine traits (vigor, architecture, resistance to pests, resistance to 
diseases, tolerance to drought, yield, marketability, taste) and overall appreciation). 
The farmers also provided their overall appreciation about the varieties, i.e., which
variety has the best and the worst performance based on the overall performance considering
all the traits.

We used the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020]. 
To consider the effect of climate on the variety performance, we used climate data
For obtaining the climate data, we used the *nasapower* package [@nasapower]. 
Climatic indices were computed with the *climatrends* package [@climatrendspkg].

First, we load the required packages and data. The ``nicabean`` dataset contains the 
tricot rankings (``trial``), the climate covariates (``covar``) and the traits observations (``traits``).
``` {r, message = FALSE, eval = TRUE, echo = TRUE}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("nasapower")
library("ggplot2")

@@

```
Make a PlackettLuce rank using the function ``rank_numeric``.
```{r, message = FALSE, eval = TRUE, echo = TRUE}
R <- vector(mode = "list", length = length(traits))

@@

```
The following lines show the first 10 rankings in the format 
used by the PlackettLuce package.  
```{r fetch, message = FALSE, eval = TRUE, echo = TRUE}
head(R[[1]], 5)
```
The correlation among rankings can be assessed using the function ``kendallTau``.  
```{r}
baseline <- which(grepl("OverallAppreciation", traits))

@@

We fit a series of Plackett-Luce models using the function ``PlackettLuce``
from the package of the same name as input in the lapply function.

@@

The ``worth_map`` function can be used to visually assess variety performance
based on different traits. The values represented in a worth_map are worth estimates
in log scale.

@@

Plackett-Luce tree using environmental data 
```

Model selection using crossvalidation
```

pseudoR2

```r

``regret``
```r

 ``reliability``

@@

We acknowledge Olga Spellman (Science Writing Service of the Alliance of
Bioversity International and CIAT) for English editing of this
manuscript
\newpage

>>>>

---
title: '`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation'
abstract: null
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

Participatory experimentation approaches have been increasingly applied in agricultural research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general
agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA]. Recently developed approaches for on-farm experimentation,
such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019]. On the other hand, newly proposed approaches for synthesis of crop variety evaluation data largely depend on the analysis of ranking data [@brown2020data].
The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce [@luce_individual_1959; @Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and Plackett-Luce models are available in R with the packages BradleyTerry2 and PlackettLuce respectively [@BradleyTerry2; @Turner2020]. However, extended functionality was required for the entire data science workflow, which usually includes: (1) Data preparation and cleaning, (2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2
respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

The R package gosset provides functionality supporting the analysis workflows in agricultural experimentation, especially with rank-based approaches. The package is available in The Comprehensive R Archive Network (CRAN) and can be installed by executing `install.packages("gosset")`. The package is named in honor to William Sealy Gosset, known by the pen name ‘Student’, a pioneer of modern statistics in small sample experimental design and analysis [@Ziliak_2019].

@@

### Data management and preparation

* `rank_binomial` transforms a ranking object into a binary comparisons, as required by package BradleyTerry2 [@BradleyTerry2].  
* `rank_numeric` converts numeric values into rankings. The parameter `ascending = ` indicates if the rankings should be made considering the numeric values in ascending order. The default is `asceding = FALSE`. This function is useful when the data were collected as numerical observations, for instance, in a experiment measuring crop yield.  
* `rank_tricot` transforms data in tricot format into rankings.

### Modelling

* `AIC` computes the Akaike Information Criterion @akaike1974] for a Bradley-Terry model fitted with BradleyTerry2 [@BradleyTerry2] or a Plackett-Luce model fitted with [@Turner2020].  
* `btpermute` deviance-based forward variable selection [@lysen2009permuted] procedure for Bradley-Terry models.
* `compare` compare the agreenment between methods.  
* `crossvalidation` performs k-fold cross-validation, where k could be specified by the user.The default is 10-fold. Folds can be provided as a vector for a custom cross-validation, such as blocked cross-validation.
* `forward` executes forward variable selection with cross-validation.  
* `kendallTau` computes the Kendall-tau rank correlation coefficient between two rankings [@kendall_1938].  
* `kendallW` computes Kendall's W (coefficient of concordance) among observed rankings and those predicted by the Plackett-Luce model.  
*   `pseudoR2` computes goodness-of-fit measure McFadden's pseudo-R2 [@mcfadden1973conditional].

### Visualization and results presentation

* `plot` provides an alternative to the default plot function of the PlackettLuce package.

* `regret`

* `reliability`

* `worth_bar`

* `worth_map`

# Illustrative Examples

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties has the best and worst performance in nine traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste) and overall appreciation). The farmers also provided their overall appreciation about the varieties, i.e., which variety has the best and the worst performance based on the overall performance considering all the traits.

We used the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020]. To consider the effect of climate on the variety performance, we used climate data. For obtaining the climate data, we used the *nasapower* package [@nasapower]. Climatic indices were computed with the *climatrends* package [@climatrendspkg].

First, we load the required packages and data. The `nicabean` dataset contains the tricot rankings (`trial`), the climate covariates (`covar`) and the traits observations (`traits`).

``` {r, message = FALSE, eval = TRUE, echo = TRUE}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("nasapower")
library("ggplot2")

@@

Make a PlackettLuce rank using the function `rank_numeric`.

```{r, message = FALSE, eval = TRUE, echo = TRUE}
R <- vector(mode = "list", length = length(traits))

@@

The following lines show the first 10 rankings in the format used by the PlackettLuce package.

```{r fetch, message = FALSE, eval = TRUE, echo = TRUE}
head(R[[1]], 5)
```

The correlation among rankings can be assessed using the function `kendallTau`.

```{r}
baseline <- which(grepl("OverallAppreciation", traits))

@@

We fit a series of Plackett-Luce models using the function `PlackettLuce` from the package of the same name as input in the lapply function.

@@

The `worth_map` function can be used to visually assess variety performance based on different traits. The values represented in a worth_map are worth estimates in log scale.

@@

@@

We acknowledge Olga Spellman (Science Writing Service of the Alliance of Bioversity International and CIAT) for English editing of this manuscript
